---
title: "Graphics with ggplot2"
author: "Bethany"
date: "22/10/2018"
output: html_document
---
#ggplot2



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
#install.packages('tidyverse')
library(tidyverse)



```

```{r}
head(ca)
View(ca)
```



```{r}
#National Parks in California
ca <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/ca.csv") 

#Acadia National Park
acadia <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/acadia.csv")

#Southeast US National Parks
se <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/se.csv")

#2016 Visitation for all Pacific West National Parks
visit_16 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/visit_16.csv")

#All Nationally designated sites in Massachusetts
mass <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/mass.csv")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

Plot visitation to all CA National Parks over time
geom_point == scatter plot
```{r}
ggplot(data=ca)+
  geom_point(aes(year, visitors))
```
```{r}
ggplot(data=ca)+
  geom_point(aes(year, visitors, colour=park_name))
```
```{r}
ggplot(data=ca)+
  geom_point(aes(year, visitors, colour=park_name))+labs(x="Year", y="visitation", "Calfifornia National Park Visitation")+theme_bw()+theme(legend.title=element_blank())
```


```{r}
head(se)
ggplot(data=se)+
  geom_point(aes(year, visitors, colour=state))+labs(x="Year", y="Visitation", "Calfifornia National Park Visitation")+theme_classic()+theme(legend.title=element_blank())+theme(axis.text.x = element_text(angle=90))
```

#Faceting plots
```{r}
ggplot(data=se)+
  geom_point(aes(year, visitors, colour=park_name))+labs(x="Year", y="Visitation", "Calfifornia National Park Visitation")+theme_bw()+theme(legend.title=element_blank())+facet_wrap(~state, scales="free")+theme(axis.text.x = element_text(angle=90))
```

#Geoms

geom_jitter
geom_violin
alpha increases the transparency of the points
colour inside aesthetics looks for a variable to make different colours, outside aesthetics colours all points one specific colour
```{r}
ggplot(data=se)+geom_jitter(aes(x=park_name, y=visitors), alpha=0.4, width= 0.1)+coord_flip()
```

```{r}
ggplot(se) + geom_line(aes(x=year, y=visitors, colour= park_name))+ theme_bw()
```
Multiple geoms on one plot
```{r}
#ggplot(acadia)+ geom_point(aes(year, visitors))+geom_line(aes(year, visitors))+geom_smooth(aes(year, visitors))

####Both creates the same plot, 2nd less redundant
ggplot(acadia, aes(year, visitors))+ geom_point()+geom_line()+geom_smooth()
```

install and library ggthemes gives you more themes


#Barcharts
Only need an x variable 
to get rid of a stacked plot use position= "dodge"
```{r}
ggplot(data=visit_16, aes(state, visitors, fill= park_name))+geom_bar(stat="identity")
```

#How to save a ggplot

Can specify the size and resolution, can save as a pdf, jpeg etc.
Can make plots interactive using,
ggplotly
```{r}
myplot<-ggplot(data=visit_16, aes(state, visitors, fill= park_name))+geom_bar(stat="identity", position= "dodge")
myplot

#to save
ggsave(filename = "myplot.jpg", myplot)
```





